Visual Interrogation of Attention-Based Models: \\ Applications in Language Inference and Machine Comprehension

### 2018-05-29 EMNLP-demo-paper/intro.tex
add perturbation discussion

Recently, attention networks have become a widely-adopted mechanism \cite{VaswaniShazeerParmar2017, Parikh2016}, the attention not only improves the model performance, but also yields interpretable intermediate representations (e.g. alignment among words). As a result, attention has been used as a natural interface for analyzing the internals of neural networks. Conduct analysis directly on the raw attention values can be challenging for human user, therefore, visual representations that highlight word alignments between sentences have been adopted, such as bipartite graph and heat map of the attention matrix~\cite{LiChenHovy2015, li2016understanding, lee2017interactive}.  However, many challenges remains. Firstly, for some NLP tasks, the word sequence pair among which the attentions are computed can be highly asymmetrical (i.e., in machine comprehension, the context paragraph can be much longer than the question sentence), where the standard visual encoding can not adequately handle.  In addition, many previous works present the attention in a statics setting. Yet, the ability to provide an interactive environment, where the user can instantaneously look at how changes in the input will affect the attention, and how small changes in attention will affect the prediction are crucial for interpreting the model. Finally, many previous visualization efforts, despite reveal many interesting results, focus more on illustrating potentially useful visualization technique than provide a flexible software tool that can be easily integrated into existing code base. As a result, applying these proposed techniques to real word examples may involving substantial engineering effort than many user may not affort to undertake.

To address these challenges in interpretating attention-based models, 
we introduce a flexible python library that allow user to easily generate a web-based interactive visual analytics environment for various usage via a composable visualization component design.
%
To facilitate exploratory analysis (see Figure~\ref{fig:modelPipeline}), we adopted a fully interactive pipeline, where the user can perturb the input text (via two perturbation schemes discussed in Section~\ref{sec:perturb}) and observe the effects on attention and prediction, or modify the attention to understand how it affect the predictions.
%
We demonstrate our system on two primary NLP tasks: natural language
inference (NLI) and machine comprehension (MC). For the NLI task, we visualize
decomposable attention networks which represents a series of strongly performing models \cite{Parikh2016}.
For the MC task, we present your visualization methods on bidirectional attention flow model,
which also represents a line of state-of-the-art models \cite{Seo2016}.

\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{pipeline}
 \caption{
 Perturbation-driven exploration of the end-to-end models that follows the encoder, attention, and classifier structure. The user can generate small perturbation of the input (i.e., paraphrasing)
 }
\label{fig:modelPipeline}
\end{figure}

>>>>

Recently, attention networks have become a widely-adopted mechanism \cite{VaswaniShazeerParmar2017, Parikh2016}, the attention not only improves the model performance, but also yields interpretable intermediate representations (e.g. alignment among words). As a result, attention has been used as a natural interface for analyzing the internals of neural networks. Conduct analysis directly on the raw attention values can be challenging for human user, therefore, visual representations that highlight word alignments between sentences have been adopted, such as bipartite graph and heat map of the attention matrix~\cite{LiChenHovy2015, li2016understanding, lee2017interactive}.  However, many challenges remains. Firstly, for some NLP tasks, the word sequence pair among which the attentions are computed can be highly asymmetrical (i.e., in machine comprehension, the context paragraph can be much longer than the question sentence), where the standard visual encoding can not adequately handle.  In addition, many previous works present the attention in a statics setting. Yet, the ability to provide an interactive environment, where the user can instantaneously look at how changes in the input will affect the attention, and how small changes in attention will affect the prediction are crucial for interpreting the model. Finally, many previous visualization efforts, despite reveal many interesting results, focus more on illustrating potentially useful visualization technique than provide a flexible software tool that can be easily integrated into existing code base. As a result, applying these proposed techniques to real word examples may involving substantial engineering effort than many user may not afford to undertake.

To address these challenges in interpreting attention-based models, 
we introduce a flexible python library that allow user to easily generate a web-based interactive visual analytics environment for various usage via a composable visualization component design.
%
To facilitate exploratory analysis (see Figure~\ref{fig:modelPipeline}), we adopted a fully interactive pipeline, where the user can perturb the input text (via two perturbation schemes discussed in Section~\ref{sec:perturb}) and observe the effects on attention and prediction, or modify the attention to understand how it affect the predictions.
%
We demonstrate our system on two primary NLP tasks: natural language
inference (NLI) and machine comprehension (MC). For the NLI task, we visualize
decomposable attention networks which represents a series of strongly performing models \cite{Parikh2016}.
For the MC task, we present your visualization methods on bidirectional attention flow model,
which also represents a line of state-of-the-art models \cite{Seo2016}.

### 2018-05-29 EMNLP-demo-paper/relatedWorks.tex
add perturbation discussion

In the recent work~\cite{LiChenHovy2015}, the composability of the vector-based representation of text is investigate by utilizing instance based attribution techniques originated from the vision community (i.e., \cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017}). 
%
In the representation erasure work~\cite{li2016understanding}, the authors explain the neural model decisions by exploring the impact of various parts of the model by altering or removing the corresponding representations (i.e., the dimension count of hidden units, or input words).

>>>>

In the recent work~\cite{LiChenHovy2015}, the composability of the vector-based representation of text is investigated by utilizing instance-based attribution techniques originated from the vision community (i.e., \cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017}). 
%
In the representation erasure work~\cite{li2016understanding}, the authors explain the neural model decisions by exploring the impact of various parts of the model by altering or removing the corresponding representations (i.e., the dimension count of hidden units, or input words).

### 2018-05-29 EMNLP-demo-paper/system.tex
add perturbation discussion

\section{System}

\subsection{Perturbation-Driven Exploration}
\label{sec:perturb}

\subsection{Attention Visualization}
As illustrated in Figure~\ref{fig:attention}(a)(b), the most widely adopted technique to bipartie graph

Attention matrix

\begin{figure*}[t]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{attentionPanels}
  \vspace{-6mm}
 \caption{
Attention visualization. In the graph attention view (a), a bipartite graph encoding is adopted, in which the edge thickness corresponds to the attention value. In the matrix attention view (b), the entries of $i^{th}$ row represent the probabilities of words in hypotheses align to the $i^{th}$ word in the premise.
The user can alter the attention values via the pop-up interface illustrated in (c).
We overlay the dependency tree ($a_1$) grammar structure to highlight important words and allow simplification of complex sentence based on the dependency tree.
%
For highly asymmetric attention relationship, we utilized a zoomable hierarchical visual representation (d).
}
\label{fig:attention}
\end{figure*}

@@

modelVis.setPredictionHook(model.predict)
modelVis.setAttentionHook(model.attention)
modelVis.setReloadModelCallback(model.reloadModel)
modelVis.setSentenceHook(gen.perturbSentence)

#open browser for the web-based interactive visualization
modelVis.show()
\end{lstlisting}

>>>>

\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{pipeline}
 \caption{
 Perturbation-driven exploration of the end-to-end models that follows the encoder, attention, and classifier structure. The user can generate small perturbation of the input (i.e., replacing synonyms, paraphrasing)
 }
\label{fig:modelPipeline}
\end{figure}

\section{Visualization System}
As illustrated in Figure~\ref{fig:modelPipeline}, many recent end-to-end NLP model follows a similar encode, attent, and classify structure, in which the attention stage provide a window for peeking into the model decision making process. However, the static attention alone likely does not tell the whole story, instead we can learn more about the model by examining the dynamic between the input, the attention, and the output. To explore the relationship between different components of the pipeline, the proposed system utilizes a perturbation-driven exploration strategy, where the  user can manually or automatically generate small changes in the input sentence and then view the aggregated prediction information. Beside the perturbation of the input, the system also allows interactive modification of the attention values, where the prediction will instantaneously updated to reflect the changes. 
%

\subsection{Input Sentences Perturbation}
\label{sec:perturb}
Due to the discrete nature of the natural language, applying perturbation to sentence for sensitivity analysis can be particularly challenging compared to other problem domain, as small changes in words can leading to drastic different in the semantics of the sentences.
To reduce the potential semantic deviation, we first devise a simple perturbation method by replacing \emph{nouns} and \emph{verbs} by their synonyms in the wordNet~\cite{Miller1995}. However, the limitation is apparent, as synonyms replacement does not guarantee the meaning of the sentence remain the same. And the wordNet also has a rather inclusive definition for synonyms, where many infrequent/obscure usages exists, which leads to less meaningful perturbed sentence. However, this method requires minimal computation and easy to code and understand.

\begin{figure}[htbp]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{sentence}
 \caption{
The interface for showing input sentences. The user can manually alter the words or apply automatic perturbation/paraphrasing of the inputs. In the \emph{perturbed} dropdown menu, the blue words are the only not exist in the original sentence.
 }
\label{fig:sentence}
\end{figure}

To improve the perturbation quality, we also adopted a translation based paraphrasing technique that is similar to the method discussed in \cite{mallinson2017paraphrasing}. Here, we translate the original English sentence into several other languages are then translate back into the English. Provided the translation produce correct result, we essential obtain paraphrasing of the original sentences (see Figure~\ref{fig:sentence}, where the drop-down menu shows the paraphrased/perturbed sentence).

\begin{figure*}[t]
\centering
\vspace{-2mm}
 \includegraphics[width=1.0\linewidth]{attentionPanels}
  \vspace{-6mm}
 \caption{
Attention visualization. In the graph attention view (a), a bipartite graph encoding is adopted, in which the edge thickness corresponds to the attention value. In the matrix attention view (b), the entries of $i^{th}$ row represent the probabilities of words in hypotheses align to the $i^{th}$ word in the premise.
The user can alter the attention values via the pop-up interface illustrated in (c).
We overlay the dependency tree ($a_1$) grammar structure to highlight important words and allow simplification of complex sentence based on the dependency tree.
%
For highly asymmetric attention relationship, we utilized a zoomable hierarchical visual representation (d).
}
\label{fig:attentionVis}
\end{figure*}

\subsection{Attention Visualization}
The system consistence several visualization components that can be combine to generate an functional interactive system. Here we will focus on the attention visualization components. 
As illustrated in Figure~\ref{fig:attentionVis}(a)(b), the most widely adopted technique to bipartie graph (a), as well as the heatmap attention matrix representation (b). 
%
In the graph attention view (Fig.~\ref{fig:attentionVis}(a)), a bipartite graph encoding is adopted, in which the edge thickness corresponds to the attention value. The color of the rectangle text block encodes the sum of all edge values connected to it (darker shade of blues correspond to higher values).
%%%%%%%%
The graph view is suitable for highlighting the most dominant alignments. However, the edges may become cluttered if multiple attention values are high. The matrix attention view (Fig.~\ref{fig:attentionVis}(b)) resolve these issues, despite being more verbose and less efficient in highlighting the dominant alignments. 
%Together, the graph and matrix views complement each other and provide the same information from different perspectives. As illustrated by the pink arrowed lines, we can see how the same attention value is visualized in both the graph and the matrix view.
To help the user recognize the correspondence during the exploration, we enable the linkage between highlighted actions in both views (see Fig.~\ref{fig:attentionVis}(a)(b), the attention of the two ``couple'' is highlighted).
%
To address the challenge of visualizing long sentence, we incorporated the grammar structure.
One augmentation

However, when the one of the text sequence become signification

@@

modelVis.setPredictionHook(model.predict)
modelVis.setAttentionHook(model.attention)
modelVis.setSentenceHook(gen.perturbSentence)

#open browser for the web-based visualization
modelVis.show()
\end{lstlisting}
